---
title: "Portfolio Correlation Heatmap"
output: html_notebook
---

# Portfolio Correlation Analysis

This notebook generates a correlation heatmap for stocks in your portfolio, showing how they move together over different time periods.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required libraries
library(quantmod)
library(jsonlite)
library(ggplot2)
library(reshape2)
library(corrplot)
library(RColorBrewer)
```

## Load Portfolio Configuration

```{r load_config}
# Load config.json
config_file <- "config.json"

if(!file.exists(config_file)) {
  stop("ERROR: config.json not found in working directory!")
}

config <- fromJSON(config_file)
tickers <- config$portfolio$stocks

# Clean and validate tickers
tickers <- toupper(trimws(tickers))
tickers <- tickers[!is.na(tickers) & tickers != ""]

cat("Portfolio loaded from config.json\n")
cat("Number of stocks:", length(tickers), "\n")
cat("Tickers:", paste(tickers, collapse = ", "), "\n")
```

## Configuration Parameters

```{r parameters}
# Analysis parameters
years_lookback <- 2  # Years of historical data
correlation_window <- 60  # Rolling correlation window (trading days)

# Calculate date range
end_date <- Sys.Date()
start_date <- end_date - (years_lookback * 365)

cat("\nAnalysis Configuration:\n")
cat("Date range:", as.character(start_date), "to", as.character(end_date), "\n")
cat("Correlation window:", correlation_window, "days\n")
```

## Fetch Stock Data

```{r fetch_data}
# Function to safely fetch stock data
get_data_safe <- function(symbol, start_date, end_date) {
  tryCatch({
    data <- getSymbols(symbol, src = "yahoo", 
                      from = start_date, to = end_date, 
                      auto.assign = FALSE, warnings = FALSE)
    return(data)
  }, error = function(e) {
    return(NULL)
  })
}

# Fetch data for all stocks
cat("\nFetching stock data...\n")
stock_data_list <- list()
valid_tickers <- character(0)

for(ticker in tickers) {
  cat(sprintf("  %s...", ticker))
  data <- get_data_safe(ticker, start_date, end_date)
  
  if(!is.null(data)) {
    stock_data_list[[ticker]] <- data
    valid_tickers <- c(valid_tickers, ticker)
    cat(" OK\n")
  } else {
    cat(" FAILED\n")
  }
}

if(length(valid_tickers) < 2) {
  stop("ERROR: Need at least 2 valid stocks for correlation analysis")
}

cat("\nSuccessfully loaded", length(valid_tickers), "stocks\n")
```

## Calculate Returns

```{r calculate_returns}
# Calculate daily log returns for each stock
returns_list <- list()

for(ticker in valid_tickers) {
  data <- stock_data_list[[ticker]]
  prices <- if(ncol(data) >= 6) Ad(data) else Cl(data)
  returns <- diff(log(prices))
  returns_list[[ticker]] <- na.omit(returns)
}

# Find common dates across all stocks
common_dates <- Reduce(intersect, lapply(returns_list, index))

cat("Common trading days:", length(common_dates), "\n")

# Align all returns to common dates
returns_matrix <- do.call(cbind, lapply(valid_tickers, function(ticker) {
  returns_list[[ticker]][common_dates]
}))

colnames(returns_matrix) <- valid_tickers

cat("Returns matrix dimensions:", nrow(returns_matrix), "x", ncol(returns_matrix), "\n")
```

## Calculate Overall Correlation Matrix

```{r overall_correlation}
# Calculate correlation matrix for entire period
correlation_matrix <- cor(returns_matrix, use = "complete.obs")

cat("\nOverall Correlation Matrix (", years_lookback, " years):\n", sep = "")
print(round(correlation_matrix, 3))

# Calculate Average Portfolio Correlation (APC)
# Average of all pairwise correlations (excluding diagonal)
upper_tri_indices <- upper.tri(correlation_matrix)
apc <- mean(correlation_matrix[upper_tri_indices])

cat("\nAverage Portfolio Correlation (APC):", round(apc, 3), "\n")
cat("Interpretation: ")
if(apc < 0.3) {
  cat("Well diversified\n")
} else if(apc < 0.6) {
  cat("Moderate diversification\n")
} else {
  cat("Concentrated exposure - stocks move together\n")
}
```

## Visualize Correlation Heatmap

```{r heatmap_plot, fig.width=10, fig.height=8}
# Create correlation heatmap using ggplot2
correlation_melted <- melt(correlation_matrix)
colnames(correlation_melted) <- c("Stock1", "Stock2", "Correlation")

# Create the heatmap
p1 <- ggplot(correlation_melted, aes(x = Stock1, y = Stock2, fill = Correlation)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = sprintf("%.2f", Correlation)), 
            color = "black", size = 3, fontface = "bold") +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, limit = c(-1, 1),
    name = "Correlation"
  ) +
  labs(
    title = "Portfolio Correlation Matrix",
    subtitle = paste0(years_lookback, "-Year Period (", 
                     format(start_date, "%Y-%m-%d"), " to ", 
                     format(end_date, "%Y-%m-%d"), ")"),
    x = "", y = "",
    caption = paste0("Average Portfolio Correlation: ", round(apc, 3))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray40"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    panel.grid = element_blank(),
    plot.caption = element_text(size = 11, face = "bold", hjust = 0.5)
  ) +
  coord_fixed()

print(p1)
```

## Identify High and Low Correlation Pairs

```{r correlation_pairs}
# Extract upper triangle (avoid duplicates)
upper_tri <- upper.tri(correlation_matrix)
correlation_pairs <- data.frame(
  Stock1 = rep(valid_tickers, each = length(valid_tickers))[upper_tri],
  Stock2 = rep(valid_tickers, length(valid_tickers))[upper_tri],
  Correlation = correlation_matrix[upper_tri]
)

# Sort by correlation
correlation_pairs <- correlation_pairs[order(-correlation_pairs$Correlation), ]

cat("\n=== Highest Correlations (Top 5) ===\n")
print(head(correlation_pairs, 10), row.names = FALSE)

# Identify high correlation pairs (> 0.7)
high_corr_pairs <- correlation_pairs[correlation_pairs$Correlation > 0.7, ]
if(nrow(high_corr_pairs) > 0) {
  cat("\n=== WARNING: High Correlation Pairs (>0.7) ===\n")
  cat("These stocks move very similarly - may represent concentrated bets:\n")
  print(high_corr_pairs, row.names = FALSE)
} else {
  cat("\n✓ No high correlation pairs (>0.7) detected\n")
}
```